---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---


```{r}
library(dplyr)
library(stringr)
library(tidytext)
library(gsubfn)
library(ggplot2)
library(ggthemes)
```


Question: Which injuries are most common in the NFL?
  
To explore this question, we will present summary statistics and graphs demonstrating injury frequency Will include specific injuries along with injuries categorized by body part (knee, head, shoulder, etc.)


Read in the data set
```{r}
injuries <- read.csv("../../Data/all_injuries_clean.csv")
```

```{r}
head(injuries)
dim(injuries)
```


Make all injury types lower case to help with string processing
```{r}
injuries <- injuries %>% mutate(injury_types = tolower(injury_types))
```



There are a lot of different levels in the injury types, so we are going to look at the most common ones (specifically) and then we will look at injuries by parts of the body.
```{r}
head(levels(as.factor(injuries$injury_types)))
```

The parts of the body will be categorized as: Head, Shoulder, Upper Torso, Lower Torso, Internal, Arm, Hand, Leg and Foot. Each of these categories has the following injury type:

Head:
  concussion
  head
  neck
  eye
  ear
  chin
  migranes
  migrane
  jaw
  nose
  tooth 
  stinger
  faciallaceration
  eyelid
  facial
  mouth
  nose
  throat
  seizure
  

Shoulder:
  shoulder
  clavicle
  collarbone
  leftshoulder
  rightshoulder
  joint shoulder
  scapula
  sternoclavicular 
  
  
Upper Torso (includes internal injuries) :
  rib
  back
  pectoral/pec
  chest
  oblique 
  lumbar
  core
  spine
  trapezius
  kidney
  lung
  heart
  abdomen
  abdominal
  solarplexus
  spleen
  hernia
  arrhythmia
  liver
  stomach
  appendix
  
Lower Torso:
  glute
  buttocks
  hip/righthip
  pelvis
  tailbone
  

  
Arm: 
  arm/rightupperarm/forearm
  bicep/biceps
  elbow/rightelbow
  tricep/triceps

  
Hand:
  hand/righthand
  thumb/rightthumb
  finger/rightfinger
  wrist/rightwrist
  
  
Leg:
  hamstring/righthamstring
  knee/rightknee/acl/mcl/meniscus/bothknees
  thigh/rightthigh
  calf/rightcalf
  quadricep/rightquadricep/quad
  groin/rightgroin
  lowerlegtibia/fibula/shin/rightshin
  adductor
  contusion

  
  
Foot:
  foot/rightfoot
  ankle
  achilles
  toe/toes
  heel





Total injuries for each injury, categorized by specific body part
```{r}
# Head
concussion <- sum(str_detect(injuries$injury_types, pattern = "concussion"))
head <- sum(str_detect(injuries$injury_types, pattern = "head"))
neck <- sum(str_detect(injuries$injury_types, pattern = "neck"))
eye <- sum(str_detect(injuries$injury_types, pattern = "eye"))
ear <- sum(str_detect(injuries$injury_types, pattern = "ear"))
chin <- sum(str_detect(injuries$injury_types, pattern = "chin"))
migraine <- sum(str_detect(injuries$injury_types, pattern = "migraine|migraines"))
jaw <- sum(str_detect(injuries$injury_types, pattern = "jaw"))
nose <- sum(str_detect(injuries$injury_types, pattern = "nose"))
tooth <- sum(str_detect(injuries$injury_types, pattern = "tooth"))
stinger <- sum(str_detect(injuries$injury_types, pattern = "stinger"))
faciallaceration <- sum(str_detect(injuries$injury_types, pattern = "faciallaceration"))
eyelid <- sum(str_detect(injuries$injury_types, pattern = "eyelid"))
facial <- sum(str_detect(injuries$injury_types, pattern = "facial"))
mouth <- sum(str_detect(injuries$injury_types, pattern = "mouth"))
nose <- sum(str_detect(injuries$injury_types, pattern = "nose"))
throat <- sum(str_detect(injuries$injury_types, pattern = "throat"))
seizure <- sum(str_detect(injuries$injury_types, pattern = "seizure"))



individual_head <- c(concussion, head, neck, eye, ear, chin, migraine, jaw, nose, tooth, stinger, faciallaceration, eyelid, facial, mouth, nose, throat, seizure)
total_head <- concussion + head + neck + eye + ear + chin + migraine + jaw + nose + tooth + stinger + faciallaceration + eyelid + facial + mouth + nose + throat + seizure

total_head
individual_head


```
We have a total of 2488 injuries for the head, with concussions being the leading injury


```{r}
# Shoulder

shoulder <- sum(str_detect(injuries$injury_types, pattern = "shoulder|leftshoulder|jointshoulder|rightshoulder"))
clavicle <- sum(str_detect(injuries$injury_types, pattern = "clavicle"))
collarbone <- sum(str_detect(injuries$injury_types, pattern = "collarbone"))
scapula <- sum(str_detect(injuries$injury_types, pattern = "scapula"))
sternoclavicular <- sum(str_detect(injuries$injury_types, pattern = "sternoclavicular"))


individual_shoulder <- c(shoulder, clavicle, collarbone, scapula, sternoclavicular)
total_shoulder <- shoulder + clavicle + collarbone + scapula + sternoclavicular


total_shoulder
individual_shoulder
```
There are a total of 1980 overall shoulder injuries, with shoulder being the greatest injury type




```{r}
# Upper Torso

rib <- sum(str_detect(injuries$injury_types, pattern = "rib|ribs"))
back <- sum(str_detect(injuries$injury_types, pattern = "back"))
pectoral <- sum(str_detect(injuries$injury_types, pattern = "pectoral|pec"))
chest <- sum(str_detect(injuries$injury_types, pattern = "chest"))
oblique <- sum(str_detect(injuries$injury_types, pattern = "oblique"))
lumbar <- sum(str_detect(injuries$injury_types, pattern = "lumbar"))
core <- sum(str_detect(injuries$injury_types, pattern = "core"))
spine <- sum(str_detect(injuries$injury_types, pattern = "spine"))
trapezius <- sum(str_detect(injuries$injury_types, pattern = "trapezius"))
kidney <- sum(str_detect(injuries$injury_types, pattern = "kidney"))
lung <- sum(str_detect(injuries$injury_types, pattern = "lung"))
heart <- sum(str_detect(injuries$injury_types, pattern = "heart"))
abdominal <- sum(str_detect(injuries$injury_types, pattern = "abdomen|abdominal"))
solarplexus <- sum(str_detect(injuries$injury_types, pattern = "solarplexus"))
spleen <- sum(str_detect(injuries$injury_types, pattern = "spleen"))
hernia <- sum(str_detect(injuries$injury_types, pattern = "hernia"))
arrhythmia <- sum(str_detect(injuries$injury_types, pattern = "arrhythmia"))
liver <- sum(str_detect(injuries$injury_types, pattern = "liver"))
stomach <- sum(str_detect(injuries$injury_types, pattern = "stomach"))
appendix <- sum(str_detect(injuries$injury_types, pattern = "appendix"))


total_uppertorso <- rib + back + pectoral + chest + oblique + lumbar + core + spine + trapezius + kidney + lung + heart + abdominal + solarplexus + spleen + hernia + arrhythmia + liver + stomach + appendix
individual_uppertorso <- c(rib, back, pectoral, chest, oblique, lumbar, core, spine, trapezius, kidney, lung, heart, abdominal, solarplexus, spleen, hernia, arrhythmia, liver, stomach, appendix)

total_uppertorso
individual_uppertorso
```
There are a total of 2,173 upper torso injuries, with back being the most frequent injury type



```{r}
# Lower Torso

glute <- sum(str_detect(injuries$injury_types, pattern = "glute|buttocks"))
hip <- sum(str_detect(injuries$injury_types, pattern = "hip|righthip|lefthip"))
pelvis <- sum(str_detect(injuries$injury_types, pattern = "pelvis"))
tailbone <- sum(str_detect(injuries$injury_types, pattern = "tailbone"))


total_lowertorso <- glute + hip + pelvis + tailbone
individual_lowertorso <- c(glute, hip, pelvis, tailbone)

total_lowertorso
individual_lowertorso
```

There are a total of 810 lower torso injuries with hip injuries being the most frequent



```{r}
# Arm

arm <- sum(str_detect(injuries$injury_types, pattern = "arm|forearm|rightupperarm"))
elbow <- sum(str_detect(injuries$injury_types, pattern = "elbow|rightelbow|leftelbow"))
tricep <- sum(str_detect(injuries$injury_types, pattern = "tricep|triceps"))
bicep <- sum(str_detect(injuries$injury_types, pattern = "bicep|biceps"))



total_arm <- arm + elbow + tricep + bicep
individual_arm <- c(arm, elbow, tricep, bicep)

total_arm
individual_arm

```
There are a total of 694 total arm injuries with elbow injuries being the most frequent





```{r}
# Hand

hand <- sum(str_detect(injuries$injury_types, pattern = "righthand|lefthand|hand"))
thumb <- sum(str_detect(injuries$injury_types, pattern = "thumb|rightthumb|leftthumb"))
finger <- sum(str_detect(injuries$injury_types, pattern = "finger|rightfinger|leftfinger"))
wrist <- sum(str_detect(injuries$injury_types, pattern = "wrist|rightwrist|leftwrist"))

total_hand <- hand + thumb + finger + wrist
individual_hand <- c(hand, thumb, finger, wrist)

total_hand
individual_hand
```

There is a total of 1003 hand injuries with overall hand injuries being the most frequent 




```{r}
# Leg

hamstring <- sum(str_detect(injuries$injury_types, pattern = "hamstring|righthamstring|lefthamstring"))
knee <- sum(str_detect(injuries$injury_types, pattern = "knee|rightknee|leftknee|acl|mcl|meniscus|bothknees"))
thigh <- sum(str_detect(injuries$injury_types, pattern = "thigh|rightthigh|leftthigh"))
calf <- sum(str_detect(injuries$injury_types, pattern = "calf|leftcalf|rightcalf"))
quadricep <- sum(str_detect(injuries$injury_types, pattern = "quadricep|leftquadricep|rightquadricep|quad"))
groin <- sum(str_detect(injuries$injury_types, pattern = "groin|rightgroin|leftgroin"))
lowerleg <- sum(str_detect(injuries$injury_types, pattern = "tibia|fibula|lowerleg|shin|rightshin|leftshin"))
adductor <- sum(str_detect(injuries$injury_types, pattern = "adductor"))
contusion <- sum(str_detect(injuries$injury_types, pattern = "contusion"))



total_leg <- hamstring + knee + thigh + calf + quadricep + groin + lowerleg + adductor + contusion
individual_leg <- c(hamstring, knee, thigh, calf, quadricep, groin, lowerleg, adductor, contusion)

total_leg
individual_leg
```
There were a total of 9212 leg injuries with knee injuries being the most frequent






```{r}
# Foot

foot <- sum(str_detect(injuries$injury_types, pattern = "foot|rightfoot|leftfoot"))
ankle <- sum(str_detect(injuries$injury_types, pattern = "ankle|rightankle|leftankle"))
achilles <- sum(str_detect(injuries$injury_types, pattern = "achilles"))
toe <- sum(str_detect(injuries$injury_types, pattern = "toe|toes"))
heel <- sum(str_detect(injuries$injury_types, pattern = "heel"))

total_foot <- foot + ankle + achilles + toe + heel
individual_foot <- c(foot, ankle, achilles, toe, heel)


total_foot
individual_foot
```

There are a total of 4942 foot injuries with ankle injuries being the most frequent



Creating a new data frame including injury types, counts and body part
```{r}
injurytype <- c("concussion", "head", "neck", "eye", 'ear', 'chin', 'migraine', 'jaw', 'nose', 'tooth', "stinger", "faciallaceration", "eyelid", "facial", "mouth", "nose", "throat", "seizure",
                "shoulder", "clavicle", "collarbone", "scapula", "sternoclavicular",
                'rib', 'back', 'pectoral', 'chest', 'oblique', 'lumbar', "core", "spine", 'trapezius', 'kidney', 'lung', 'heart', 'abdominal', 'solarplexus', 'spleen', 'hernia', "arrhythmia", "liver", "stomach", "appendix",
                'glute', 'hip', 'pelvis', "tailbone",
                'arm', 'elbow', 'tricep', "bicep",
                'hand', 'thumb', 'finger', 'wrist',
                'hamstring', 'knee', 'thigh', 'calf', 'quadricep',  'groin', 'lowerleg', 'adductor', "contusion",
                'foot', 'ankle', 'achilles', 'toe', 'heel')

count <- c(concussion, head, neck, eye, ear, chin, migraine, jaw, nose, tooth, stinger, faciallaceration, eyelid, facial, mouth, nose, throat, seizure, 
           shoulder, clavicle, collarbone, scapula, sternoclavicular,
           rib, back, pectoral, chest, oblique, lumbar, core, spine, trapezius, kidney, lung, heart, abdominal, solarplexus, spleen, hernia, arrhythmia, liver, stomach, appendix,
           glute, hip, pelvis, tailbone,
           arm, elbow, tricep, bicep,
           hand, thumb, finger, wrist,
           hamstring, knee, thigh, calf, quadricep, groin, lowerleg, adductor, contusion,
           foot, ankle, achilles, toe, heel)

bodypart <- c(rep("Head", length(individual_head)), 
              rep("Shoulder", length(individual_shoulder)), 
              rep("Upper Torso", length(individual_uppertorso)),
              rep("Lower Torso", length(individual_lowertorso)),
              rep("Arm", length(individual_arm)),
              rep("Hand", length(individual_hand)),
              rep("Leg", length(individual_leg)),
              rep("Foot", length(individual_foot))
              )

injury_count <- data.frame(injurytype, count, bodypart)

injury_count %>% arrange(desc(count))

```



Let's look at the distribution of each specific injury
```{r, fig.height=9}
injury_count %>%
  mutate(injurytype = reorder_within(injurytype,count, bodypart)) %>%
  ggplot(aes(y=injurytype, x=count)) +
  geom_col()+
  scale_y_reordered() +
  geom_text(aes(label = count), position = position_dodge(width = 0.9), vjust = .4, hjust = -.1, fontface = 'bold') +
  xlab("Count") +
  ylab("Injury Type") + 
  ggtitle("Distribution of Specific Injury Types") +
  theme_economist() +
  theme(
    axis.title.x = element_text(size = 14, vjust = -3),
    axis.title.y = element_text(size = 14, vjust = 3),
    title = element_text(size = 18),
    plot.title = element_text(hjust = 0.5, vjust = 2.5)
  )

```


We see that injuries towards the bottom are not that frequent. Let's group the injuries together based on where they are in the body and look at a more general distribution instead. (See Willow_Lauren.EDA)


                                                                