---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidytext)
library(gsubfn)
library(tidyr)
library(ggthemes)
```

Read in the data set
```{r}
setwd("C:/Users/wduff/OneDrive/School/Harvard/Fall2021/BST260/BST_260_project")

injury <- read.csv("Data/all_injuries_clean.csv")
players = read.csv("Data/all_player_demographic_clean.csv")

# Merge data
injuries = left_join(injury, players, by = c("name", "team", "year", "full_team"))
```

```{r}
head(injuries)
dim(injuries)
```


```{r}
injuries[11:18]
```

Gathering into long format
```{r}
injury_gather <- gather(injuries, key = "bodypart", value = "counts", 11:18)
injury_gather
```

## EDA

Overall distribution of body part injuries
```{r}
injury_gather %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() + 
  geom_text(aes(label = counts), position = position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Injuries by Body Part") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Now for the next part at EDA, let's look at the injury distributions across the various positions in football
```{r}
levels(as.factor(injuries$position_id))
```

```{r}
injuries %>% filter(position_id == "K")
```



We see we have the following positions: Kicker (K), Offensive Line (OL), Punter (P), Quarter Back (QB), Running Back (RB), Tight End (TE), Wide Reciever (WR) and Defense (DEF).

Since Defense has it's own category with no specific position (like Linebacker, Defensive Line or Safety), let's first compare the injury distributions between Offensive Players and Defensive Players

```{r}
offensive_position <- c("K", "OL", "P", "QB", "RB", "TE", "WR")

# Only offensive players
offense <- injury_gather %>% filter(position_id %in% offensive_position)
# Only defensive players
defense <- injury_gather %>% filter(position_id == "DEF") 

levels(as.factor(offense$position_id))
levels(as.factor(defense$position_id))

dim(offense)
dim(defense)
```

We have 7712 offensive players and 72309 defensive players which is great since the data sets are somewhat balanced and therefore comparing them will be valid. 


Overall distribution of offensive injuries
```{r}
offense %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() +
  geom_text(aes(label = counts), position = position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Offensive Injuries") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Overall distribution of defensive injuries
```{r}
defense %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() +
  geom_text(aes(label = counts), position = position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Defensive Injuries") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Distribution of injuries for kickers
```{r}
offense %>% 
  filter(position_id == "K") %>%
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_text(aes(label = counts), position=position_dodge(width = 0.9), vjust = -0.25, fontface='bold') +
  geom_col() +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distribution of Offensive Injuries") +
  theme_economist() +
  theme(
    axis.title.x = element_text(size = 14, vjust = -3),
    axis.title.y = element_text(size = 14, vjust = 3),
    title = element_text(size = 18),
    plot.title = element_text(hjust = 0.5)
  )
```






