---
title: "EDA for P1"
author: "Addison McGhee"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F)
```


```{r, include=F}
library(AER)
library(caret)
library(e1071)
library(eeptools)
library(foreign)
library(gam)
library(ggrepel)
library(ggthemes)
library(gridExtra)
library(lubridate)
library(magrittr)
library(nnet)
library(pander)
library(pROC)
library(pscl)
library(randomForest)
library(rpart)
library(splines)
library(splines2)
library(splitstackshape)
library(tidyverse)
library(VGAM)
library(vcd)
library(MASS)
options(digits = 3)
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r}
# load data
injured = read.csv("../../Data/all_injuries_clean.csv") 
demographics = read.csv("../../Data/all_player_demographic_clean.csv")

# add column for total number of injuries for each player each year
injured <- injured %>%
  mutate(all_injuries = select(injured, c("head", "shoulder", "upper_torso", "lower_torso", 
                                           "arm", "hand", "leg", "foot")) %>% rowSums)

# join with demographic data
injuries = full_join(injured, demographics, by = c("name", "full_team", "team", "year"))

# replace NA injuries with zero? double-check that I should do this
injuries$all_injuries[is.na(injuries$all_injuries)] <- 0
```

**clarifying question--does the injuries dataset only include people who were injured? need to make sure that we are sure about who was not injured**


```{r, include=F}
fix_nfl_names <- function(x){ #https://rdrr.io/github/papagorgio23/bettoR/src/R/fix_nfl_names.R
  x[grep("Arizona Cardinals", x, ignore.case=TRUE)] <- "ARI"
  
  x[grep("Atlanta Falcons", x, ignore.case=TRUE)] <- "ATL"

  x[grep("Baltimore Ravens", x, ignore.case=TRUE)] <- "BAL"

  x[grep("Buffalo Bills", x, ignore.case=TRUE)] <- "BUF"

  x[grep("Carolina Panthers", x, ignore.case=TRUE)] <- "CAR"

  x[grep("Chicago Bears", x, ignore.case=TRUE)] <- "CHI"

  x[grep("Cincinnati Bengals", x, ignore.case=TRUE)] <- "CIN"

  x[grep("Cleveland Browns", x, ignore.case=TRUE)] <- "CLE"

  x[grep("Dallas Cowboys", x, ignore.case=TRUE)] <- "DAL"

  x[grep("Denver Broncos", x, ignore.case=TRUE)] <- "DEN"

  x[grep("Detroit Lions", x, ignore.case=TRUE)] <- "DET"

  x[grep("Green Bay Packers", x, ignore.case=TRUE)] <- "GB"

  x[grep("Houston Texans", x, ignore.case=TRUE)] <- "HOU"

  x[grep("Indianapolis Colts", x, ignore.case=TRUE)] <- "IND"

  x[grep("Jacksonville Jaguars", x, ignore.case=TRUE)] <- "JAX"

  x[grep("Kansas City Chiefs", x, ignore.case=TRUE)] <- "KC"

  x[grep("Miami Dolphins", x, ignore.case=TRUE)] <- "MIA"

  x[grep("Minnesota Vikings", x, ignore.case=TRUE)] <- "MIN"

  x[grep("New England Patriots", x, ignore.case=TRUE)] <- "NE"

  x[grep("New Orleans Saints", x, ignore.case=TRUE)] <- "NO"

  x[grep("New York Jets", x, ignore.case=TRUE)] <- "NYJ"

  x[grep("New York Giants", x, ignore.case=TRUE)] <- "NYG"

  x[grep("Las Vegas Raiders", x, ignore.case=TRUE)] <- "OAK"

  x[grep("Philadelphia Eagles", x, ignore.case=TRUE)] <- "PHI"

  x[grep("Pittsburgh Steelers", x, ignore.case=TRUE)] <- "PIT"

  x[grep("Los Angeles Chargers", x, ignore.case=TRUE)] <- "LAC"

  x[grep("Los Angeles Rams", x, ignore.case=TRUE)] <- "LAR"

  x[grep("San Francisco 49ers", x, ignore.case=TRUE)] <- "SF"

  x[grep("Seattle Seahawks", x, ignore.case=TRUE)] <- "SEA"

  x[grep("Tampa Bay Buccaneers", x, ignore.case=TRUE)] <- "TB"

  x[grep("Tennessee Titans", x, ignore.case=TRUE)] <- "TEN"

  x[grep("Washington Redskins", x, ignore.case=TRUE)] <- "WAS"

  return(x)
}

age <- function(dob, age.day = today(), units = "years", floor = TRUE) {
    calc.age = interval(dob, age.day) / duration(num = 1, units = units)
    if (floor) return(as.integer(floor(calc.age)))
    return(calc.age)
}
```


```{r, include=F}
injuries = injuries %>% mutate(short_team_name = fix_nfl_names(full_team),
                               age = age(ymd(birthdate))) %>% 
                        #select(name, full_team, age,
                               #year, games_in_season, 
                               #num_games_missing, num_games_injured, injury_types,
                               #earliest_injury, latest_injury, position_id, 
                               #height_inches, weight_pounds, birthdate,
                               #short_team_name) %>% 
                        filter(position_id != "")
```


```{r}

table(injuries$position_id)
```

&nbsp;

# P1

### Q1: Which positions are most at-risk for injuries? The table below contains descriptive statistics calculated for each position over multiple seasons (2009 - Present). 

```{r}
# number_injury = injuries %>% group_by(position_id, year) %>% 
#                              count()
# 
# number_injury %>% filter(position_id != "WR") %>% 
#   group_by(position_id) %>% 
#   summarise(mean = mean(n), 
#             sd = sd(n), 
#             Q1 = quantile(n, probs = 0.25),
#             median = median(n),
#             Q3 = quantile(n, probs = 0.75)) %>% 
#   pander


# count injuries per position per year
number_injury <- injuries %>%
  group_by(year, position_id) %>%
  summarize(n = sum(all_injuries, na.rm = TRUE))

pander(number_injury)

number_injury %>% filter(position_id != "WR") %>% 
  group_by(position_id) %>% 
  summarise(mean = mean(n), 
            sd = sd(n), 
            Q1 = quantile(n, probs = 0.25),
            median = median(n),
            Q3 = quantile(n, probs = 0.75)) %>%
  pander()

# try the same thing but count multiple injuries multiple times
```                                                   

**why take out WR?**


```{r}             
number_injury1 %>% ggplot(aes(year, n, color = position_id)) +
                    geom_line() +
                    xlab("Year") +
                    ylab("# of Injuries") +
                    scale_y_log10() +
                    scale_color_discrete(name = "Position") +
                    ggtitle("Number of Injuries by Position (2009 - Present)")
```

**need to adjust each position somehow**

```{r}
# adjust for numbers of players in that position we have in the data?

# adjust for typical number of people in that position on the field at one time?


```



&nbsp;

### What about injuries per year? Is this uniformly distributed?
```{r}
injuries %>% filter(year != 2021) %>%
              ggplot(aes(year)) + 
              geom_bar() + 
              ylab("# of Injuries") +
              ggtitle("# of Players Injured per Season")
```

**confused by this plot**

&nbsp;

### We now will switch to another dataset that contains information for both injured and non-injured players. 
* For the next analyses, we will only look at the **2018** season
* We will perform logistic regression, kNN, and random forest using the binary outcome `injury`. 
* Our predictors will be continuous bmi and the categorical variable `position_id`. 
* We will then compare the results of the different methods based on accuracy, sensitivity/specificity, and ROC/AUC.


&nbsp;

```{r}
#players_dat = read.csv("all_player_demographic_clean.csv")

players = injuries %>% 
            filter(year == 2018) %>% 
            mutate(injury = as.factor(ifelse(all_injuries == 0, 0, 1)), # make binary injury variable
                   position_id = as.factor(position_id), # make position categorical
                   bmi = weight_pounds / height_inches^2, # make BMI variable
                   age = age(ymd(birthdate)) - 3) %>% # subtract 3 years to match age in 2018
            filter(position_id != "")
```

**tried removing NAs--need to check how many there were**

&nbsp;

### Our first step is to check for missing data in our predictors and outcome
```{r}
apply(players %>% 
        select(injury, bmi, height_inches, weight_pounds, position_id, age), 2, 
      function(x) sum(is.na(x))) %>% pander
```

&nbsp;

**Fortunately, there are no missing observations for these features in the 2018 season.**

&nbsp;

### What is the distribution of injuries by position in 2018?

```{r}
players$position_id = 
  fct_relevel(players$position_id, c("DEF", "OL", "WR", "RB", "TE", "QB", "K", "P", ""))

players %>% 
  ggplot(aes(position_id, fill = injury)) +
    geom_bar(position = "dodge") +
    xlab("Position") +
    ylab("# of Players") +
    ggtitle("Injured Players by Position in 2018") +
    scale_fill_manual(name = "Injury", guide = guide_legend(reverse=TRUE), values = c("#619CFF", "#F8766D"))
  
```

&nbsp;

**The graph above is very useful for determining whether a logistic regression coefficient would likely be statistically significant. We are looking to see, for a given position, if there is a large discrepancy between the number of injured and non-injured players. If the number of players injured is much larger than the number of non-injured players, then we would expect that the odds of getting injured in this position is large. We would expect the opposite to be the case for a position where the number of non-injured players is much greater than the number of injured players.**

&nbsp;

### What is the distribution of BMI for the injured vs non-injured players?
```{r}
players %>% ggplot(aes(injury, bmi)) +
              geom_boxplot()
```

&nbsp;

**The absence of any significant difference between groups suggests that `bmi` will not be a significiant predictor in our Logistic model.**

&nbsp;

### What is the distribution of `injury` by `age` in 2018?
```{r}
injuries %>% filter(year == 2018) %>% 
             mutate(age = age - 3) %>% # subtract 3 years for 2018 age
             ggplot(aes(age)) +
               geom_histogram(binwidth = 0.5) +
               xlab("Age") +
               ylab("# of Injuries") +
               ggtitle("# of Injuries by Age in 2018") +
               scale_x_continuous(breaks = seq(0, 50, 1))
```

&nbsp;

## Try side by side bars for `age`
```{r}
players %>% ggplot(aes(age, fill = injury)) +
              geom_histogram(binwidth = 0.5, position = "dodge") +
              xlab("Age") +
              ylab("# of Injuries") +
              ggtitle("# of Injuries by Age in 2018") +
              scale_x_continuous(breaks = seq(0, 50, 1)) +
              scale_fill_manual(name = "Injury", 
                                guide = guide_legend(reverse=TRUE), values = c("#619CFF", "#F8766D"))
  
```

&nbsp;

## Try scaling by number of players in each age group
```{r}
age_group_size = players %>% group_by(age) %>% count() # get size of each age group

injured_age_group_size = players %>% # get size of age group for injured players only
                          filter(injury == 1) %>% 
                          group_by(age) %>% 
                          count()

# Some age groups have no injuries, so the vector lengths differ
group_size = age_group_size %>% filter(age %in% injured_age_group_size$age) %>% .$n

injured_age_group_size$scaled_n = injured_age_group_size$n / group_size # Scale by group size
head(injured_age_group_size)

injured_age_group_size %>% ggplot(aes(age, scaled_n)) +
                            geom_col() +
                            xlab("Age") +
                            ylab("Percentage") +
                            ggtitle("Percentage of Players Injured by Age Group") +
                            scale_x_continuous(breaks = seq(0, 50, 1)) +
                            scale_fill_manual(name = "Injury", 
                              guide = guide_legend(reverse=TRUE), values = c("#619CFF", "#F8766D"))
  
```

&nbsp;


## Try quadratic age
```{r}

```

&nbsp;

### Logistic regression (model only; check for significant coefficients)

```{r}
players = players %>% mutate(position_id = relevel(position_id, ref = "TE"))

logi_fit = glm(data = players, injury ~ position_id + bmi + age, family = binomial())
pander(summary(logi_fit))

```

&nbsp;

### Why is `DEF` not significant? We see there is almost an even split between injured and non-injured players
```{r}
players %>% filter(injury == 1) %>% select(position_id) %>% table() %>% pander(caption = "Injured")
players %>% filter(injury == 0) %>% select(position_id) %>% table() %>% pander(caption = "non-Injured")
```

## Try stepwise selection

### First remove missing values
**need to check how many missing vals--add to Addison's table**

```{r}
players_cc <- players %>%
  drop_na(age, game_starter, NFL_draft_round, NFL_draft_year, position_id, weight_pounds, height_inches, bmi)
```

**if missing NFL draft round, replace with some higher number? if they are worse players**

```{r}
# model with all potential covariates
# logi_fit_all <- glm(data = players, injury ~ as.factor(position_id) + age + NFL_draft_round + 
#                       as.factor(game_starter) + weight_pounds + height_inches + bmi + NFL_draft_year, 
#                     family = binomial())
logi_fit_all <- glm(data = players_cc, injury ~ 1, 
                    family = binomial())
summary(logi_fit_all)

logi_step_fit_for <- logi_fit_all %>% 
  step(direction = "forward", data = players_cc, trace = TRUE, 
       scope = (~ as.factor(position_id) + age + NFL_draft_round + as.factor(game_starter) + weight_pounds +
                  height_inches + bmi + NFL_draft_year))
summary(logi_step_fit_for)


# scope = (~ as.factor(position_id) + age + NFL_draft_round + as.factor(game_starter) + weight_pounds +
#                   height_inches + bmi + NFL_draft_year)

logi_step_fit_back <- logi_fit_all %>% 
  step(direction = "backward", data = players_cc, trace = TRUE, 
       scope = (~ as.factor(position_id) + age + NFL_draft_round + as.factor(game_starter) + weight_pounds +
                  height_inches + bmi + NFL_draft_year))
summary(logi_step_fit_back)
```

**some predictors have missing values**

&nbsp;

# Machine Learning and Prediction

&nbsp;

### Create training and test set

```{r}
set.seed(1)

x <- stratified(players, "injury", 0.7, keep.rownames = TRUE)
train_set <- x %>% dplyr::select(-rn)
train_index <- as.numeric(x$rn)
test_set <- players[-train_index,]

dim(train_set)
dim(test_set)
```

&nbsp;

#### NOTE: Accuracy improved after removing `bmi` as a predictor

### Logistic regression (using `position_id` and `age` as features)

```{r, logistic regression}
logi_fit = glm(data = train_set, injury ~ position_id + age, family = binomial())

p_hat_logit = predict(logi_fit, newdata = test_set, type = "response") # get predicted probabilities

y_hat_logit <- as.factor(ifelse(p_hat_logit > 0.5, 1, 0)) # set threshold of 0.5

confusionMatrix(data = y_hat_logit, 
                reference = test_set[["injury"]], positive = "1") # generate confusion matrix

```




&nbsp;



&nbsp;

### k-Nearest Neighbors

#### Use 2-fold cross-validation to find optimal `k` (achieves the highest accuracy for 0.5 cut-off)
```{r}
set.seed(1)
control <- trainControl(method = 'cv', number = 2, p = .5) # cut-off of 0.5

players2 <- mutate(players, label = injury) %>%
  select(label, position_id, age)

res <- train(label ~ .,
             data = players2,
             method = "knn",
             trControl = control,
             tuneLength = 1, # How fine a mesh to go on grid
             tuneGrid = data.frame(k = seq(3, 151, 2)),
             metric = "Accuracy")

res$bestTune # best choice of 'k' is k = 131
plot(res)
```

```{r kNN}
knn_fit = knn3(data = train_set, injury ~ position_id + age, k = 131) # fit kNN using optimal k = 131

f_hat_knn = predict(knn_fit, newdata = test_set)[ , 2] # get probability for injury

y_hat_knn = as.factor(ifelse(f_hat_knn > 0.5, 1, 0))

confusionMatrix(data = y_hat_knn, 
                reference = test_set[["injury"]], positive = "1") # generate confusion matrix

```

&nbsp;

### Random Forest

```{r Random Forest}
rf_fit = randomForest(injury ~ position_id + age, 
                      data = players[train_index, ]) # fit random forest using training set

rf_fit

p_hat_rf = predict(rf_fit, newdata = test_set, type = "prob")[,2] # get predicted probabilities

y_hat_rf = as.factor(ifelse(p_hat_rf > 0.5, 1, 0)) # use 0.5 cutoff to get estimates

confusionMatrix(data = y_hat_rf, 
                reference = test_set[["injury"]], positive = "1") # generate confusion matrix

```

&nbsp;

&nbsp;

```{r, message=FALSE}
roc_logi = roc(test_set[["injury"]], p_hat_logit) # ROC curve creation
roc_knn = roc(test_set[["injury"]], f_hat_knn)
roc_rf = roc(test_set[["injury"]], p_hat_rf)

ggroc(list("Logistic Regression" = roc_logi, "kNN, k = 11" = roc_knn, "Random Forest" = roc_rf)) +
  theme(legend.title = element_blank()) +
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "black", linetype = "dashed") +
  xlab("Sensitivity") +
  ylab("Specificity") 

```


```{r}
auc(roc_logi) # calculate AUC for each ROC curve
auc(roc_knn)
auc(roc_rf)
```

&nbsp;



## Multinomial Outcome

**We will now return to the original `injuries` dataset. The goal now will be to now categorize the injury types to a new factor with multiple levels. This type of outcome is often used for multinomial regression. To maintain independence of observations, we will only examine injuries from the 2018 season.**

```{r}
'injuries_2018 = injuries %>% filter(year == 2018) %>% 
                      mutate(multi_injury = 
                               case_when(
                                 
                                )
                               )'
```





