---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidytext)
library(gsubfn)
library(tidyr)
library(ggthemes)
```

Read in the data set
```{r}
injuries <- read.csv("all_injuries_clean.csv")
```

```{r}
head(injuries)
dim(injuries)
```

```{r}
# Change the injuries to be all lower case
injuries <- injuries %>% mutate(injury_types = tolower(injury_types))
```

```{r}
# Replace any "/" with " " ?
injuries$injury_types <- gsub("/", " ", injuries$injury_types)
```


Create the body part vectors
```{r}
Head <- c("concussion", "head", "neck", "eye", "ear", "chin", "migranes", "migrane", "jaw", "nose", "tooth", "stinger", "faciallaceration", "eyelid", "facial", "mouth", "nose", "throat", "seizure")

Shoulder <- c("shoulder", "clavicle", "collarbone", "leftshoulder", "rightshoulder", "jointshoulder", "scapula", "sternoclavicular")

UpperTorso <- c("rib", "back", "pectoral", "pec", "chest", "oblique", "lumbar", "core", "spine", "trapezius", "kidney", "lung", "heart", "abdomen", "abdominal", "solarplexus", "spleen", "hernia", "arrhythmia", "liver", "stomach", "appendix")

LowerTorso <- c("glute", "buttocks", "hip", "righthip", "pelvis", "tailbone")


Arm <- c("arm", "rightupperarm", "forearm", "bicep", "biceps", "elbow", "rightelbow", "tricep", "triceps")

Hand <- c("hand", "righthand", "thumb", "rightthumb", "finger", "rightfinger", "wrist", "rightwrist")

Leg <- c("hamstring", "righthamstring", "knee", "rightknee", "acl", "mcl", "meniscus", "bothknees", "thigh", "rightthigh", "calf", "rightcalf", "quadricep", "rightquadricep", "quad", "groin", "rightgroin", "lowerleg", "tibia", "fibula", "shin", "rightshin", "adductor", "contusion")

Foot <- c("foot", "rightfoot", "ankle", "achilles", "toe", "toes", "heel")
```


Function to count broad body part injuries 
```{r}
# Bodypart input must be a string
broad <- function(bodylist, bodypart) {
  
  pasted <- paste(bodylist, collapse = "|")
  broad_injury <- gsub(pasted, bodypart, injuries$injury_types)
  str_count(broad_injury, pattern = bodypart)

  }
```



Calculating broad injury counts
```{r}
# Head
head_counts <- broad(Head, "head")

# Shoulder
shoulder_counts <- broad(Shoulder, "shoulder")

# Upper Torso
uppertorso_counts <- broad(UpperTorso, "uppertorso")

# Lower Torso
lowertorso_counts <- broad(LowerTorso, "lowertorso")

# Arm
arm_counts <- broad(Arm, "arm")

# Hand
hand_counts <- broad(Hand, "hand")

# Leg
leg_counts <- broad(Leg, "leg")

# Foot
foot_counts <- broad(Foot, "foot")
```


Adding body part columns to the orginal injury data frame
```{r}
injuries <- injuries %>% 
  mutate(head = head_counts) %>%
  mutate(shoulder = shoulder_counts) %>%
  mutate(upper_torso = uppertorso_counts) %>%
  mutate(lower_torso = lowertorso_counts) %>%
  mutate(arm = arm_counts) %>%
  mutate(hand = hand_counts) %>%
  mutate(leg = leg_counts) %>%
  mutate(foot = foot_counts)
  

head(injuries)
```


Save the cleaned data frame
```{r}
write.csv(injuries,"C:/Users/wduff/OneDrive/School/Harvard/Fall2021/BST260/BST_260_project/cleaneddata.csv", row.names = TRUE)
```


Gathering into long format
```{r}
injury_gather <- gather(injuries, key = "bodypart", value = "counts", 28:ncol(injuries))
injury_gather
```

## EDA

Overall distribution of body part injuries
```{r}
injury_gather %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() + 
  geom_text(aes(label = counts), position= position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Injuries by Body Part") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Now for the next part at EDA, let's look at the injury distributions across the various positions in football
```{r}
levels(as.factor(injuries$position_id))
```

```{r}
injuries %>% filter(position_id == "K")
```



We see we have the following positions: Kicker (K), Offensive Line (OL), Punter (P), Quarter Back (QB), Running Back (RB), Tight End (TE), Wide Reciever (WR) and Defense (DEF).

Since Defense has it's own category with no specific position (like Linebacker, Defensive Line or Safety), let's first compare the injury distributions between Offensive Players and Defensive Players

```{r}
offensive_position <- c("K", "OL", "P", "QB", "RB", "TE", "WR")

# Only offensive players
offense <- injury_gather %>% filter(position_id %in% offensive_position)
# Only defensive players
defense <- injury_gather %>% filter(position_id == "DEF") 

levels(as.factor(offense$position_id))
levels(as.factor(defense$position_id))

dim(offense)
dim(defense)
```

We have 7712 offensive players and 72309 defensive players which is great since the data sets are somewhat balanced and therefore comparing them will be valid. 


Overall distribution of offensive injuries
```{r}
offense %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() +
  geom_text(aes(label = counts), position= position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Offensive Injuries") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Overall distribution of defensive injuries
```{r}
defense %>% 
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_col() +
  geom_text(aes(label = counts), position= position_dodge(width = 0.9), vjust = -0.25, fontface = 'bold') +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distriubtion of Defensive Injuries") +
  theme_economist() +
  theme(axis.title.x = element_text(size = 16, vjust = -3),
        axis.title.y = element_text(size = 16, vjust = 3),
        title = element_text(size = 20),
        plot.title = element_text(hjust = 0.5)
                )
```



Distribution of injuries for kickers
```{r}
offense %>% 
  filter(position_id == "K") %>%
  group_by(bodypart) %>%
  summarise(counts = sum(counts)) %>%
  ggplot(aes(x = reorder(bodypart, -counts), y = counts)) +
  geom_text(aes(label = counts), position=position_dodge(width = 0.9), vjust = -0.25, fontface='bold') +
  geom_col() +
  xlab("Body Part") +
  ylab("Count") +
  ggtitle("Distribution of Offensive Injuries") +
  theme_economist() +
  theme(
    axis.title.x = element_text(size = 14, vjust = -3),
    axis.title.y = element_text(size = 14, vjust = 3),
    title = element_text(size = 18),
    plot.title = element_text(hjust = 0.5)
  )
```






