---
title: "260 Project EDA"
author: "Group Name: K-Nearest Tailgaters"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, message = F)
```


```{r, include=F}
library(foreign)
library(gam)
library(ggrepel)
library(ggthemes)
library(gridExtra)
library(magrittr)
library(nnet)
library(pander)
library(splines2)
library(splines)
library(tidyverse)
library(VGAM)
options(digits=3)
theme_update(plot.title = element_text(hjust = 0.5))
```

```{r}
injuries = read.csv("injuries.csv")
nfl_roster = read.csv("nfl_roster.csv")
```

&nbsp;

### Check for duplicates, NA's in outcome
```{r}
any(duplicated(injuries[["Full_Name"]])) # no duplicate names (multiple injuries during the year)
any(is.na(injuries[["Injury"]])) # no NA's in outcome
```

&nbsp;

### `Injury.Status`: What to do with non-football injuries (NFI-R), or COVID? How can we sort based on injury severity?
```{r}
table(injuries$Injury.Status)
```

&nbsp;

### One idea: IR > Doubtful > Questionable, then remove NFI and COVID
```{r}
var_change = function(x) {
  ordinal_injury = c()
  for(i in seq_along(x)) {
    if (str_detect(x[i], "NFI") | str_detect(x[i], "Non Football Injury")) {
      ordinal_injury[i] = "NFI"
    }
    else if (str_detect(x[i], "Questionable")) {
      ordinal_injury[i] = "Questionable"
    }
    else if (str_detect(x[i], "Doubtful")) {
      ordinal_injury[i] = "Doubtful"
    }
    else if (str_detect(x[i], "COVID-19")) {
      ordinal_injury[i] = "COVID-19"
    }
    else if (str_detect(x[i], "IR") | str_detect(x[i], "Physically Unable to Perform")) {
      ordinal_injury[i] = "IR"
    }
    else {
      ordinal_injury[i] = "other"
    }
  }
  ordinal_injury
}

injuries %<>% mutate(ordinal_injury = var_change(injuries$Injury.Status))
table(injuries$ordinal_injury)

```

&nbsp;

### What are the "other" injuries?
```{r}
injuries %>% filter(ordinal_injury == "other") %>% select(Injury, Injury.Status) # Not the kneecap!!
```

&nbsp;

### Maybe "Out for Week X" == "IR"?
```{r}
injuries$ordinal_injury[injuries$ordinal_injury == "other"] = "IR"
```

&nbsp;

### Well, it seems that players are really only "Questionable" or on "IR".

### We can remove "COVID-19" and "NFI" injuries, and merge "Doubtful" with "Questionable":
```{r}
injuries %<>% filter(!ordinal_injury %in% c("COVID-19", "NFI")) %>% 
              mutate(binary_injury = ifelse(ordinal_injury == "IR", 1, 0))
```

&nbsp;

```{r, include=F} 
fix_nfl_names <- function(x){ #https://rdrr.io/github/papagorgio23/bettoR/src/R/fix_nfl_names.R
  x[grep("Arizona Cardinals", x, ignore.case=TRUE)] <- "ARI"
  
  x[grep("Atlanta Falcons", x, ignore.case=TRUE)] <- "ATL"

  x[grep("Baltimore Ravens", x, ignore.case=TRUE)] <- "BAL"

  x[grep("Buffalo Bills", x, ignore.case=TRUE)] <- "BUF"

  x[grep("Carolina Panthers", x, ignore.case=TRUE)] <- "CAR"

  x[grep("Chicago Bears", x, ignore.case=TRUE)] <- "CHI"

  x[grep("Cincinnati Bengals", x, ignore.case=TRUE)] <- "CIN"

  x[grep("Cleveland Browns", x, ignore.case=TRUE)] <- "CLE"

  x[grep("Dallas Cowboys", x, ignore.case=TRUE)] <- "DAL"

  x[grep("Denver Broncos", x, ignore.case=TRUE)] <- "DEN"

  x[grep("Detroit Lions", x, ignore.case=TRUE)] <- "DET"

  x[grep("Green Bay-Packers", x, ignore.case=TRUE)] <- "GB"

  x[grep("Houston Texans", x, ignore.case=TRUE)] <- "HOU"

  x[grep("Indianapolis Colts", x, ignore.case=TRUE)] <- "IND"

  x[grep("Jacksonville Jaguars", x, ignore.case=TRUE)] <- "JAX"

  x[grep("Kansas City-Chiefs", x, ignore.case=TRUE)] <- "KC"

  x[grep("Miami Dolphins", x, ignore.case=TRUE)] <- "MIA"

  x[grep("Minnesota Vikings", x, ignore.case=TRUE)] <- "MIN"

  x[grep("New England-Patriots", x, ignore.case=TRUE)] <- "NE"

  x[grep("New Orleans-Saints", x, ignore.case=TRUE)] <- "NO"

  x[grep("New York-Jets", x, ignore.case=TRUE)] <- "NYJ"

  x[grep("New York-Giants", x, ignore.case=TRUE)] <- "NYG"

  x[grep("Las Vegas-Raiders", x, ignore.case=TRUE)] <- "OAK"

  x[grep("Philadelphia Eagles", x, ignore.case=TRUE)] <- "PHI"

  x[grep("Pittsburgh Steelers", x, ignore.case=TRUE)] <- "PIT"

  x[grep("Los Angeles-Chargers", x, ignore.case=TRUE)] <- "LAC"

  x[grep("Los Angeles-Rams", x, ignore.case=TRUE)] <- "LAR"

  x[grep("San Francisco-49ers", x, ignore.case=TRUE)] <- "SF"

  x[grep("Seattle Seahawks", x, ignore.case=TRUE)] <- "SEA"

  x[grep("Tampa Bay-Buccaneers", x, ignore.case=TRUE)] <- "TB"

  x[grep("Tennessee Titans", x, ignore.case=TRUE)] <- "TEN"

  x[grep("Washington Redskins", x, ignore.case=TRUE)] <- "WAS"

  return(x)
}
```


### Merge data.frames by "Full_Name"
```{r}
# Some people who are injured are no longer on the roster => out for season
nfl_roster %<>% 
  mutate(Team = fix_nfl_names(Team)) %>% 
  select(-c(X, College, Drafted, Height, Number, Birthday, 
            Draft.Round, Draft.Pick, Birthday_string, Number))

injured_still_on_team = nfl_roster %>% 
            inner_join(injuries[, -1], by = c("Full_Name", "Short_Name", "Team"))

roster_with_injuries = nfl_roster %>% 
            left_join(injuries[,-1], by = c("Full_Name", "Short_Name", "Team")) %>% 
            mutate(ordinal_injury = 
                     case_when(str_detect(ordinal_injury, "IR") ~ 2, 
                               str_detect(ordinal_injury, "Questionable") ~ 1,
                               str_detect(ordinal_injury, "Doubtful") ~ 1,
                               is.na(ordinal_injury) ~ 0),
                   binary_injury = 
                     ifelse(is.na(binary_injury), 0, 1)
                   )

```

&nbsp;

### NA's per variable
```{r}
apply(injured_still_on_team, 2 , function(x) sum(is.na(x))) # distribution of NA's


apply(roster_with_injuries, 2 , function(x) sum(is.na(x))) 
```

&nbsp;

### Collinearity for Height/Weight
```{r}
cor.test(roster_with_injuries$height_inches, 
         roster_with_injuries$Weight) # expected, so will combine to BMI
```

&nbsp;

### Create factors and new variables of interest
```{r}
injured_still_on_team %<>% mutate(Injury = as.factor(Injury), 
                     Pos = as.factor(Pos), 
                     Team = as.factor(Team))

roster_with_injuries %<>% mutate(Pos = as.factor(Pos), 
                     Team = as.factor(Team))


Offensive_Player = c("QB", "RB", "FB", "TB", "HB", "OL", "G", "LG", "RG", 
                     "T", "LT", "RT", "C", "WR", "TE")

Defensive_Player = c("DL", "DE", "LE", "RE", "DT", "NT", "LB", "MLB", "ILB", 
                     "OLB", "LOLB", "ROLB", "DB", "CB", "S", "SS", "FS")

Special_Teams = c("P", "K", "PR")


injured_still_on_team %<>% 
  mutate(Offense = ifelse(Pos %in% c(Offensive_Player, Special_Teams), 1, 0), # offense yes/no
           BMI = (Weight / height_inches^2) * 703) # BMI

roster_with_injuries %<>% 
  mutate(Offense = ifelse(Pos %in% c(Offensive_Player, Special_Teams), 1, 0), # offense yes/no
           BMI = (Weight / height_inches^2) * 703) # BMI
```

&nbsp;

### Descriptive stats
```{r}
number_injury = injured_still_on_team %>% group_by(Team) %>% 
  summarize(num_injury = length(Injury))

pander(summary(number_injury$num_injury)) # ~10 injuries per team
```

&nbsp;

### Histogram + Boxplot of Injuries per Team
```{r}
number_injury %>%  # histogram (looks kinda bimodal...why??)
  ggplot(aes(num_injury)) + 
    geom_histogram(binwidth = 1) + 
    xlab("# of Injuries per Team")

number_injury %>% # boxplot (looks kinda symmetric)
  ggplot(aes(num_injury)) + 
    geom_boxplot() + 
    xlab("# of Injuries")
```

&nbsp;

### Injuries by Team and Position
```{r}
injured_still_on_team %>% group_by(Injury) %>% 
  summarize(num_injury = length(Injury)) %>%
  mutate(Injury = fct_reorder(Injury, num_injury, .desc = T)) %>%
  ggplot(aes(Injury, num_injury)) + 
  geom_col() +
  ylab("Count") +
  ggtitle("Distribution of Injuries") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Injuries by Team
injured_still_on_team %>% group_by(Team) %>% 
  summarize(num_injury = length(Injury)) %>% 
  arrange(desc(num_injury)) %>% 
  mutate(Team = fct_reorder(Team, num_injury, .desc = T)) %>% 
  ggplot(aes(Team, num_injury)) + geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("# of Injuries") +
  ggtitle("Injury by Team")


# Injuries by Position
injured_still_on_team %>% group_by(Pos) %>% 
  summarize(num_injury = length(Injury)) %>% 
  arrange(desc(num_injury)) %>% 
  mutate(Pos = fct_reorder(Pos, num_injury, .desc = T)) %>% 
  ggplot(aes(Pos, num_injury)) + 
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("# of Injuries") +
  xlab("Position") +
  ggtitle("Injury by Position")

```

&nbsp;

### Logistic Regression
```{r}
logi_fit = glm(data = roster_with_injuries, 
               binary_injury ~ Age + Exp. + BMI + Offense, family = "binomial")

pander(summary(logi_fit))
```

&nbsp;

### Logistic Regression w/Team + Position
```{r}
logi_fit_Team_Pos = glm(data = roster_with_injuries, 
                        binary_injury ~ Team + Pos, family = "binomial")

pander(summary(logi_fit_Team_Pos))

```

&nbsp;

**Maybe too much information in the outcome is lost by making injury binary, maybe ordinal or multinomial would be preferred.**

**We also might want to use data from past years as we are only half way through the current season.**

&nbsp;

### Multinomial
```{r}
multinom_injury = multinom(data = roster_with_injuries,
                           ordinal_injury ~ Exp.) # Experience only
summary(multinom_injury)

ggplot(roster_with_injuries, aes(Exp., multinom_injury$fitted.values[,1])) +
  geom_line(aes(Exp., multinom_injury$fitted.values[,1], color = "No Injury")) +
  geom_line(aes(Exp., multinom_injury$fitted.values[,2], color = "Questionable/Doubtful")) +
  geom_line(aes(Exp., multinom_injury$fitted.values[,3], color = "Injured Reserve")) +
  scale_color_discrete("Status") +
  ylab("Predicted Probabilities") +
  xlab("Experience")
```








